---
title: "Week 1"
pagetitle: "Week 1 BSTA 526"
description: "Intro to R/RStudio, Functions, Vectors, Data Types"
date: "`r library(here); source(here('schedule_class_dates.R')); w1d1`"
date-modified: "today"
categories: [""]
format: 
  html:
    link-external-newwindow: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source(here::here("schedule_class_dates.R"))
```


## Topics

-   Part 0: Welcome
    -   Introduction to course & expectations (syllabus)
-   Part 1: 
    -   Intro to R & RStudio
        * [Handout](../resources/R info.pdf) with directions on installing R & RStudio
    - Introduction to Quarto
    - Functions, Vectors, Data Types


## Announcements

* Class materials for BSTA 526 will be provided in the shared OneDrive folder [BSTA_526_W26_class_materials_public](https://ohsuitg-my.sharepoint.com/my?id=%2Fpersonal%2Fniederha%5Fohsu%5Fedu%2FDocuments%2Fteaching%2FBSTA%20526%2FW26%2FBSTA%5F526%5FW26%5Fclass%5Fmaterials%5Fpublic).
* For today's class, make sure to download to your computer the folder called `part1`. Open RStudio by double-clicking on the project file called `BSTA_526_W26_class_materials_public.Rproj` in the main OneDrive folder.



## Class materials


| Part | OneDrive folder | Slides | Webpage |
|--------------|--------------|--------------|--------------------------------|
| 0 | [{{< iconify logos:microsoft-onedrive >}}](https://ohsuitg-my.sharepoint.com/my?id=%2Fpersonal%2Fniederha%5Fohsu%5Fedu%2FDocuments%2Fteaching%2FBSTA%20526%2FW26%2FBSTA%5F526%5FW26%5Fclass%5Fmaterials%5Fpublic%2Fpart0) | [{{< iconify simple-icons:quarto >}}](../part0/part_00_b526_slides.qmd) |  |
| 1 | [{{< iconify logos:microsoft-onedrive >}}](https://ohsuitg-my.sharepoint.com/my?id=%2Fpersonal%2Fniederha%5Fohsu%5Fedu%2FDocuments%2Fteaching%2FBSTA%20526%2FW26%2FBSTA%5F526%5FW26%5Fclass%5Fmaterials%5Fpublic%2Fpart1) | [{{< iconify simple-icons:quarto >}}](../part1/part_01_b526_slides.qmd) | [{{< iconify streamline web size=20px >}}](../part1/part_01_b526.qmd) |

: {.striped .hover tbl-colwidths="\[3, 4, 4, 4\]"}


## Readings


### Required

These readings are mostly supplemental and will help you if there are concepts that are unclear in class.

- [R and RStudio Basics (Chapter 3)](https://ismayc.github.io/rbasics-book/3-rstudiobasics.html) - make sure to watch the videos.
    * Skip the following sections in Chapter 3:
        * 3.3.1, 3.3.3, 
- [Intro to R & Rstudio, and Quarto](https://niederhausen.github.io/BSTA_511_F25/slides_md/Day01_bsta511_md.html) from Day 1 of my BSTA 511/611 class in Fall 2025 ([slides version](https://niederhausen.github.io/BSTA_511_F25/slides/Day01_bsta511.html#/title-slide))
- [Vectors and Data Frames (Section 3.3)](https://datasciencebook.ca/wrangling.html#data-frames-vectors-and-lists) from [Data Science: A First Introduction](https://datasciencebook.ca/) by Timbers, Campbell, & Lee
- [Workflow: scripts and projects](https://r4ds.hadley.nz/workflow-scripts.html) from R for Data Science (2e) by Wickham, Cetinkaya-Rundel, & Grolemund
    * [Section 6.1: Scripts](https://r4ds.hadley.nz/workflow-scripts.html#scripts)
    * [Section 6.2: Projects](https://r4ds.hadley.nz/workflow-scripts.html#projects)


### Optional 

#### Posit/R Cheatsheets

- Many useful information and tips can be found in the list of [cheatsheets](https://posit.co/resources/cheatsheets/) on the Posit website

#### Markdown / Quarto 

*   Slides on how to do [Markdown and other formatting in Quarto files](https://ohsu-octri-berd.github.io/Quarto_BERD_2025/slides/part2_intro.html#/types-of-quarto-content-text). This is optional, but may be a helpful reference as you create your own Quarto files. 
    *   *Slides are from the OCTRI-BERD workshop on [Creating Professional Presentations and Websites using R/Quarto](https://ohsu-octri-berd.github.io/Quarto_BERD_2025/) presented by Jessica Minnier & myself in July 2025. *
    *   *There are also slides for [Creating Presentations](https://ohsu-octri-berd.github.io/Quarto_BERD_2025/slides/part3_presentations.html#/title-slide) and [Creating Websites](https://ohsu-octri-berd.github.io/Quarto_BERD_2025/slides/part4_websites.html#/title-slide) using Quarto.*
*   Quarto's website for [Markdown Basics](https://quarto.org/docs/authoring/markdown-basics.html)
    *   [Quarto YAML for html](https://quarto.org/docs/reference/formats/html.html)
    *   [Quarto front matter](https://quarto.org/docs/authoring/front-matter.html)


#### Swirl Basics

This is a resource for learning basic R concepts using the R package `swirl`. 

To start it, run the following code in the *console* in RStudio:

```{r}
#| eval: false
# first install the swirl package
library(swirl)
swirl()

# Note: type bye() to exit swirl
```

You'll want to take a look at the `R Programming` course, especially the following sections:

1. Basic Building Blocks
3. Sequences of Numbers
4. Vectors
5. Missing values


## Post-class survey

* Please fill out the [post-class survey](https://ohsu.ca1.qualtrics.com/jfe/form/SV_2f0SKdTYUo80aRo) to provide feedback. Thank you!
* Previous muddiest points and clearest points with responses are collected  [here](../survey_feedback_previous_years.qmd).

## Homework

* See OneDrive folder for homework assignment.
* HW 1 due on `r format(w2d1, "%m/%d")`.

## Recording

* __In-class__ recording links are on [Sakai](https://sakai.ohsu.edu/portal/site/BSTA-526-1-21029-W26/tool/0a1dc228-11f3-4ae0-a5d1-1d4597c01180/ShowPage?returnView=&studentItemId=0&backPath=&bltiAppStores=false&errorMessage=&clearAttr=&messageId=&source=&title=&newTopLevel=false&sendingPage=232002&postedComment=false&addBefore=&itemId=5509813&path=push&reviewAssessment=false&topicId=&addTool=-1&recheck=&id=&forumId=). Navigate to _Course Materials_ -> _Schedule with links to in-class recordings_. 





