---
title: "8: Functions and Lists"
pagetitle: "8: Functions and Lists"
subtitle: "BSTA 526: R Programming for Health Data Science"
author: "Meike Niederhausen, PhD & Jessica Minnier, PhD"
institute: "OHSU-PSU School of Public Health"
date: "2/26/2026"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 5
    theme: minty
    embed-resources: true
    html-math-method: mathjax
  # revealjs:
  #     incremental: false
  #     scrollable: true
  #     chalkboard: true
  #     theme: [../sky_modified_smaller_font_B526_W26_v2.scss]
  #     width:  1100 #1200 # 1050 #default 1050; ipad 3:5, 1600
  #     # height: 825 #900 #800 #default 700; 788 for 3:5, 1200
  #     slide-number: true
  #     number-sections: true
  #     number-depth: 5
  #     link-external-newwindow: true
  #     html-math-method: mathjax
  #     auto-stretch: false  # stop autosizing figures
execute:
  echo: true
  message: false
  warning: false
  fig.align: "center"
  fig-width: 5
  fig-height: 3
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
# The first chunk is special 
# because if you try to run a chunk below without running anything else first, 
# R will automatically run this first

knitr::opts_chunk$set(echo = TRUE, fig.path = "output_figs/")

# Here I am loading libraries/packages and installing packages you don't have yet. Add the argument `update=TRUE` if you want to update your packages.
pacman::p_load(
  vembedr,      # youtube embedder
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  readxl,       # import Excel data
  gtsummary,    # summary statistics and tests
  ggthemes,     # lots of themes
  here,         # helps with file management
  janitor,      # for data cleaning, making tables
  palmerpenguins, # penguin data
  ghibli,       # studio ghibli palettes
  paletteer,    # additional ggplot2 palettes
  gt,           # beautiful tables
  rstatix,      # summary stats for numeric vars
  naniar        # missing values exploration
  )

# set ggplot theme for slides 
# theme_set(theme_bw(base_size = 22))
# theme_update(text = element_text(size=22))  # set global text size for ggplots

```




# Welcome to R Programming: Part 8!

Today we will learn more about customizing tables, adding and reordering `factor`s, and `ggplot`.

<br>

Before you get started:

>Remember to save this notebook under a new name, such as `part_08_b526_YOURNAME.qmd`.

<br>

-   Load the packages in the setup code chunk

## Learning Objectives

-   **Learn** how to write and use custom functions
-   **Learn** about lists as a general purpose data structure
-   **Learn** and **utilize** list properties
-   **Access** list elements using `$` and `[[]]`
-   **Understand** the difference between homogeneous and heterogeneous lists
-   **Use** `purrr::pluck()` to access list elements
-   **Understand** how `data.frames()` are list-like



## Load data sets

```{r}
# load various data sets
data(penguins)

smoke_complete <- read_csv(here::here("part8", "data", "clinical.csv"))
lusc_data <- read_csv(here::here("part8", "data", "tumor", "LUSC.csv"))
meso_data <- read_csv(here::here("part8", "data", "tumor", "MESO.csv"))
prad_data <- read_csv(here::here("part8", "data", "tumor", "PRAD.csv"))
read_data <- read_csv(here::here("part8", "data", "tumor", "READ.csv"))
```


# Motivation

Today: building blocks for truly automating your data wrangling or analysis work. 

<br>

This is the true power of learning how to program!


## Custom Functions & Lists

At first these two sections will seem disjointed:

-   Functions
    -   Create your own functions
    -   Use for data wrangling or analyses or visualization, anything
-   Lists
    -   An object type in R that can be very useful
    -   Spoiler alert: data frames can actually behave like a `list`

<br>

-   Next week we will use them together with "Functional Programming Tools" 
    -   such as `purrr::map()` type functions to iterate a function 
    -   over a list or over rows of a data frame, for example.

## Custom Functions

***Please read the [Functions](https://r4ds.hadley.nz/functions.html) chapter in R4DS for some background!*** (Chapter 25)

We know what a function is in R, for instance, these are "built-in" functions in R

```{r}
mean(1:10)
toupper("abc")
```

But what if we want to create our own custom functions?

We have actually done this before! When we used `across()`:

```{r}
penguins %>% summarize(across(.cols = ends_with("mm"),
                              .fns = ~ mean(.x, na.rm = TRUE))) # this is a custom function, in a way
```

## Motivating Example

-   Here is another example for motivating today's work.

::: {.panel-tabset}
### One dataset

-   Let's say we want some summary measures from our data, 
    -   specifically the means of days to death, age at diagnosis, and 
    -   the correlation of those two variables.

```{r}
lusc_data %>% summarize(
  mean_cig = mean(days_to_death, na.rm = TRUE),
  mean_age = mean(age_at_diagnosis, na.rm = TRUE),
  cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                    use = "complete.obs"))
```

### Many datasets

-   Now, we have read in multiple data sets, and want to calculate this *for each data set*. 
-   We can do this separately, with copy and paste and 
    -   hope we don't make a mistake, and 
    -   also keep track of all our outputs somehow:

```{r}
lusc_data %>% summarize(
  mean_cig = mean(days_to_death, na.rm = TRUE),
  mean_age = mean(age_at_diagnosis, na.rm = TRUE),
  cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                    use = "complete.obs"))

meso_data %>% summarize(
  mean_cig = mean(days_to_death, na.rm = TRUE),
  mean_age = mean(age_at_diagnosis, na.rm = TRUE),
  cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                    use = "complete.obs"))

prad_data %>% summarize(
  mean_cig = mean(days_to_death, na.rm = TRUE),
  mean_age = mean(age_at_diagnosis, na.rm = TRUE),
  cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                    use = "complete.obs"))

read_data %>% summarize(
  mean_cig = mean(days_to_death, na.rm = TRUE),
  mean_age = mean(age_at_diagnosis, na.rm = TRUE),
  cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                    use = "complete.obs"))
```

### Custom function

-   Or, we can create our own function to do this! 
-   We will go through the parts of a function below, 
    -   but this is what it would look like:

```{r}
# custom function
myfun <- function(mydata){
  mydata %>% summarize(
    mean_cig = mean(days_to_death, na.rm = TRUE),
    mean_age = mean(age_at_diagnosis, na.rm = TRUE),
    cor_cig_age = cor(days_to_death, age_at_diagnosis, 
                      use = "complete.obs"))
}

# apply custom function to each of the datasets 
# (note that the output is not being saved below)
myfun(lusc_data)
myfun(meso_data)
myfun(prad_data)
myfun(read_data)
```


### With lists

-   Later we will see that we can make a list of data frames, and 
    -   use `purrr::map()` to iterate over this list and 
    -   obtain all of the outputs at once. 
-   This is a motivating example, it will take some time to get to this point!

```{r}
list_of_data <- list("LUSC" = lusc_data, 
                     "MESO" = meso_data, 
                     "PRAD" = prad_data, 
                     "READ" = read_data)
```

::: {.panel-tabset}
#### Return a list of outputs

```{r}
# returns a list of outputs
map(list_of_data, myfun)
```

#### Return a data frame of outputs:

```{r}
# returns a data frame of outputs
map_df(list_of_data, myfun, .id = "cancer_type")
```


[*We'll start with learning about how to create our own functions.*]{style="color:purple"}

:::
:::



# Functions: Don't Repeat Yourself (DRY)

## Motivation

-   If you do something in R once and have to do it again, 
    -   it's probably best to think of how to do it *well* once, 
    -   rather than cutting and pasting again.

-   Part of doing this is by using and writing *functions*. 
    -   We can take code that we might have to write 10 times in a row, and 
    -   package it in a way that we can use it multiple times.

-   For the example above, 
    -   we have some computations we want to perform on specific variables in a data set, and 
    -   we have multiple data sets with those variables we need to do this with. 
    -   A function can be used to perform these computations.

## Advantages over using copy-and-paste

From [R for Data Science](https://r4ds.hadley.nz/functions.html):

> Writing a function has four big advantages over using copy-and-paste:
>
> 1.  You can give a function an evocative name that makes your code easier to understand.
> 2.  As requirements change, you only need to update code in one place, instead of many.
> 3.  You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).
> 4.  It makes it easier to reuse work from project-to-project, increasing your productivity over time.


## Creating a function

Let's learn about how we define a function.

```{r}
#| eval: false

f <- function(<arguments>) {
  ## Do something interesting or useful
  ## Code
  ## Return an output

}
```

Things to ask yourself:

1.    What do you want your function to do?
1.    What will the *output* of the function be? 
1.    What kind of *object* do you want the output to be?

## Example of creating a function

```{r}
square_number <- function(number) {
  output_number <- number * number 
  return(output_number)
}

square_number(4)
```

::: {.panel-tabset}
### Name & arguments

-   We define our function by 
    -   giving it a name, `square_number`, and 
    -   defining its *arguments* within the parentheses of `function()`. 

-   Our example function only has one argument, `number`:

>`square_number <- function(number) {`

### The `{`

-   What about the `{`? 
    -   It defines the beginning of a *code block* - 
        -   a code block defines the code that will run when we use the function.

### Stuff between the `{....}`

-   Between the `{....}` we calculate the square of our number and 
    -   assign it to `output_number`.
-   The next line uses `return()` to return a value from our function. 
-   Then we *close the code block with a `}`*.

### `return()`

-   The return value of a function is the last expression in the function body to be evaluated. 
    -   i.e. `return()` should appear in the last line of the function.

-   **Note that you don't *have* to use `return()`** 
    -   but it makes things clearer once your functions get complicated, and 
    -   especially if you are returning multiple outputs. 
    
### Function without `return()`   


-   The function below without `return() does the same as above.

[*If you don't specify the return object with `return()`, R will assume you meant to return the last line of code inside the function that returns a value.*]{style="color:green"}

```{r}
square_number_again <- function(number) {
  number * number
}

square_number_again(4)
```

-   However, [be careful]{style="color:darkorange"}! Notice that the code below doesn't work. 
-   Why? Because nothing was returned, 
    -   the last line does not return a value but saves an object. 
-   This is why it's best to just use `return()` 
    -   so you can be sure you're returning what you want.

```{r}
square_number_again <- function(number) {
  mynumsq <- number * number
}

square_number_again(4)

# note the lack of output when running this function!
```

:::

## Mini-challenge (on your own)

-   Make a function called `cube_number` that returns the cube of a number. 
-   Test it out and make sure it works.

-   *Hint: you can find the cube of a number using `num^3`.*

```{r}
#| eval: false
cube_number <- ___



```

## Arguments

-   We also can input multiple arguments to a function, which can be of different types:

::: {.panel-tabset}
### Example 1

```{r}
abs_fun <- function(x, i) {
  res <- abs(x - mean(x)) ^ i
  return(res)
}

```

-   What is the output of this function? 
-   What type of data object is the output?

```{r}
abs_fun(x = 1:10, i = 4)
```


### Example 2


```{r}
mypet <- function(name, pet) {
  mystring <- paste(name,"'s ", pet, sep = "")
  mystring <- toupper(mystring)
  return(mystring)
}
```

-   You can name arguments, or use them in order, just like built-in R functions

```{r}
# not naming arguments and hoping they are ordered correctly:
mypet("meike", "cat")

# oops
mypet("cat", "meike")

# naming arguments:
mypet(name = "meike",
      pet = "cat")

# naming arguments not in order:
mypet(pet = "cat", 
      name = "meike")
```

-   What if we only give one argument?

```{r}
#| eval: false

mypet("meike")
```

### Default arguments

-   We can also set default arguments. 
-   If we define an argument inside function(), 
    -   that is the default value and it no longer needs to be specified.

```{r}
mypet <- function(name, pet = "Mouse") {
  mystring <- paste(name,"'s ", pet, sep = "")
  mystring <- toupper(mystring)
  return(mystring)
}

mypet("meike", "cat")

mypet("meike")

```

:::

## Save the output of the function

-   We can save the output of a function just as we do with built-in functions:

```{r}
petstring <- mypet(name = "meike", pet = "cat")
petstring
```

-   We can also manipulate the output, since it is an object:

```{r}
tolower(petstring)

str_replace_all(string = petstring, 
                pattern = "CAT", 
                replacement = "DOG")
```



## Challenge 1 (10 min)

1.  Create a function that calculates the mean of a numeric vector, removing `NA`'s. Make sure to test that it works. What happens when you input a character vector?

2.  Create a function that takes a data frame and calculates the mean of all numeric variables using your custom mean function above. Try this with the penguins data set.

```{r}

```

## What does the function name return?

-   Type/run `cor` in the console (no parentheses). 
    -   What do you see? 
-   Also try a name of a function you created, such as `square_number`.

```{r}
# note: what yo see in the console is different than in the html output
square_number
```

## Example: standardizing values in a vector

-   Below we create our own function to *standardize* values in a vector:
    -   subtract the mean of the values and 
    -   divide by the standard deviation of the values

```{r}
# argument is a numeric vector
standardize_fun <- function(vec) {
  vec2 <- (vec - mean(vec, na.rm = TRUE)) / sd(vec, na.rm = TRUE)
  return(vec2) # returns another numeric vector
}

# test 1
tmpvec <- 1:10
standardize_fun(tmpvec)

# test 2
tmpvec <- c(-3, -2, -1, 0, 1, 2, 3)
standardize_fun(tmpvec)
```

-   Usually we create functions to do more complex tasks than simple number or text manipulation.
    -   Often the input argument is a data frame or tibble.

## Example & Mini challenge: counting categories in a variable from a data frame

1.   What are the input and output of the function below? 
1.   What requirements do we need for the input data frame? 

::: {.panel-tabset}
### Function code

```{r}
count_categories <- function(df){
  counts <- df %>%
    count(disease)
  
  return(counts)
}

```

### Apply the function

-   Apply the function to different data sets:

```{r}
# these data are loaded in the first section above

count_categories(smoke_complete)

count_categories(lusc_data)

count_categories(meso_data)

```

-   Why does this not work?
```{r}
#| eval: false

count_categories(penguins)

```


### Comments

-   This function might be useful if you needed to summarize the diseases across a large number of files.

<br>

-   One thing you might notice is that `disease` is ["hard coded"]{style="color:green"} 
    -   which means it always counts the column with that exact name. 
-   There is a way to make this more flexible 
    -   (i.e. take an argument that specifies the column name) 
    -   using something called `tidyeval`, which we will learn about next time.

:::


## Example: read in data from a file path

-   If you find yourself reading in many data sets the same way, 
-   you might create your own custom function to do this the way you like.

```{r}
# custom function to read in a file (specified with argument path)
# output is a data frame

load_files <- function(path){
  out_frame <- readxl::read_excel(path, na="NA", sheet=1)
  out_frame <- janitor::clean_names(out_frame) %>%
    mutate(path_name = path)
  return(out_frame)
}

```

-   Run the function:

```{r}
smoke1 <- load_files(path = here::here("part8", "data", "smoke_1.xlsx"))
glimpse(smoke1)            # notice last column
```

-   You don't have to include "`path =`"

```{r}
smoke2 <- load_files(here::here("part8", "data", "smoke_2.xlsx"))
glimpse(smoke2)            # notice last column
```


## Using custom functions within `mutate()` and `across()`

::: {.panel-tabset}
### Within `mutate()`

-   We can use custom functions within `mutate()`:

```{r}
# standardize_fun is a function we created above

# using inside mutate without across:
smoke1 %>% 
  mutate(age_std = standardize_fun(age_at_diagnosis)) %>% 
  # selecting just a few columns to see new column:
  select(primary_diagnosis, age_at_diagnosis, age_std)


```

### Within `mutate()` and `across()`

-   When using `across()` within the `mutate()`, we can use the custom function
-   Note that the code below replaces values in *all* numeric columns with standardized transformations of the values

```{r}
smoke1 %>% 
  mutate(across(.cols = where(is.numeric),
                .fns = standardize_fun))

# we don't need the .fns = ~ standarize_fun(.x) 
# because it is a name of a function and we aren't changing arguments
```

:::

## Creating a function inside `across()`

-   We have seen examples where we created a function inside `across()` 
-   This is a bit messier but does the same thing
-   We have to use `.x` as our function argument placeholder/name.

```{r}
smoke1 %>% 
  mutate(across(.cols = where(is.numeric),
                .fns = ~ (.x - mean(.x, na.rm = TRUE)) / sd(.x, na.rm = TRUE)
                ))


```

-   You might think of the function like the one created in the code below, 
    -   which is the same as what we specified above,
    -   just with `.x` instead of `vec`, and no `return()`

```{r}
standardize_fun <- function(.x) {
  (.x - mean(.x, na.rm = TRUE)) / sd(.x, na.rm = TRUE)
}
```


## Using custom functions within `summarize()`

-   Below we create a function and use it within `summarize()`. 
    -   Input: a numeric vector (such as a numeric column of a dataset)
    -   Output: it divides the mean of the vector by the standard deviation of the vector
-   This custom function is a "summarizing" function that returns *one* value, 
    -   *not a vector of multiple values*. 
 
```{r}
mean_over_sd <- function(vec) {
  mean(vec, na.rm = TRUE) / sd(vec, na.rm = TRUE)
}
```

-   Use the custom function within `summarize()`:

```{r}
penguins %>% 
  group_by(species) %>%
  summarize(std = mean_over_sd(bill_length_mm))

```

-   Use the custom function within `summarize()` and `across()`:
    
```{r}
penguins %>% 
  group_by(species) %>%
  summarize(across(
    .cols = where(is.numeric), 
    .fns = mean_over_sd))
```



## Challenge 2 (15 min)

1.  Use the above functions `standardize_fun()` and `mean_over_sd()` with the LUSC data to standardize the numeric vectors `age_at_diagnosis` and `cigarettes_per_day`. Calculate mean/sd for those vectors as well.

2.  Create a custom function that converts mm to cm, and apply the function within a `mutate()` and `across()` on all the columns in `penguins` data that end with "mm".

```{r}

```



## Checking for user errors with `if()`

::: {.panel-tabset}

### Motivation 

-   What happens when someone tries `square_number("two")`? 
-   Since "two" is a character, you can't multiply it. 


```{r}
#| eval: false
# run this yourself, eval: false so we can knit
square_number("two")
```

-   We can let the the user to know that this isn't what the function is expecting for its input.

### `if(){ stop() }` code

-   We can use `if(){ }` as a way to check if the input value  `number` is actually numeric.
-   `if()` statements check the condition within the parentheses,
    -   and then run the code in the `{ }` (curly brackets) if the condition is true.


```{r}
#| eval: false
# run this yourself, eval: false so we can knit

square_number_improved <- function(number) {
  # check for whether number is numeric:
  if(!is.numeric(number)) { 
    # stop gives the error and exits the function:
    stop("Your input was not a number")
  }
  
  output_number <- number * number 
  
  return(output_number)
  
}

square_number_improved(2)

square_number_improved("two")
```

### Do you need to check for all errors?

-   It depends on who uses your functions. 
-   If the functions are just for you and your lab, as long as everyone understands what goes into the functions, you probably don't need to anticipate all kinds of errors, just a few major ones.
-   It's when you are releasing your software as a freely available package that you should think about doing extensive testing for errors, especially the functions you anticipate will be used by most people who use your software.

:::



## Challenge 3 (10 minutes)

::: {.panel-tabset}

### Instructions

-   Create a custom function that does the same data cleaning steps as the code below for one dataset, so that you don't repeat the same code 3 times. Put your solution in the code chunk below this, labeled `challenge1solns`.
-   Your custom function should essentially be a copy of the code that **does these steps**:
    -   Loads the time point data from an .xlsx file, 
    -   cleans the data, 
    -   renames any necessary columns, and 
    -   then pivots the data into a long format.
-   The function should have **two arguments**: 
    -   the name of the excel file and 
    -   the name of the sheet
-   The function should output a long dataset
-   Bind the data at the end of the code chunk *after* running the function three times.
-   If you have time, add in an error check that when the sheet number is not between 2 and 4, returns an error message to say "Incorrect sheet value for time point data."


### Code to "functionize"

```{r}
#| label: challenge1

# Load data
mouse_tp1 <- read_excel(here::here("part8", "data", "mouse_biomarker.xlsx"), 
                        sheet = 2, 
                        na = c("","/"),
                        n_max = 34, # CAREFUL WITH THIS
                        .name_repair = janitor::make_clean_names) %>%
  remove_empty(which = c("rows","cols"))

mouse_tp2 <- read_excel(here::here("part8", "data", "mouse_biomarker.xlsx"), 
                        sheet = 3, 
                        na = c("","/"),
                        .name_repair = janitor::make_clean_names) %>%
  remove_empty(which = c("rows","cols"))

mouse_tp3 <- read_excel(here::here("part8", "data", "mouse_biomarker.xlsx"), 
                        sheet = 4, 
                        na = c("","/"),
                        .name_repair = janitor::make_clean_names) %>%
  remove_empty(which = c("rows","cols"))

# rename columns (do this to all data, just in case)
colnames(mouse_tp2) <- str_replace(colnames(mouse_tp2), "_ng", "_pg")

mouse_tp1 <- mouse_tp1 %>%
  rename(mirna1 = mirna_1,
         mirna2 = mi_rna_2,
         preference_obj1 = preference,
         preference_obj2 = x_4) %>%
  filter(!is.na(sid)) %>%
  mutate(
    preference_obj1 = as.numeric(preference_obj1),
    preference_obj2 = as.numeric(preference_obj2)
  ) 

mouse_tp2 <- mouse_tp2 %>%
  rename(mirna1 = mirna_1,
         mirna2 = mi_rna_2,
         preference_obj1 = preference,
         preference_obj2 = x_4) %>%
  filter(!is.na(sid)) %>%
  mutate(
    preference_obj1 = as.numeric(preference_obj1),
    preference_obj2 = as.numeric(preference_obj2)
  )  

mouse_tp3 <- mouse_tp3 %>%
  rename(mirna1 = mirna_1,
         mirna2 = mi_rna_2,
         preference_obj1 = preference,
         preference_obj2 = x_4) %>%
  filter(!is.na(sid)) %>%
  mutate(
    preference_obj1 = as.numeric(preference_obj1),
    preference_obj2 = as.numeric(preference_obj2)
  ) 

# Pivot data
mouse_tp1 <- mouse_tp1 %>%
  pivot_longer(cols = starts_with("normalized"),
               names_to = "biomarker_type",
               values_to = "biomarker_value")
mouse_tp2 <- mouse_tp2 %>%
  pivot_longer(cols = starts_with("normalized"),
               names_to = "biomarker_type",
               values_to = "biomarker_value")
mouse_tp3 <- mouse_tp3 %>%
  pivot_longer(cols = starts_with("normalized"),
               names_to = "biomarker_type",
               values_to = "biomarker_value")

# Note: the code below is not a part of your custom funcion

# Bind data
mouse_tp <- bind_rows("tp1" = mouse_tp1, 
                      "tp2" = mouse_tp2,
                      "tp3" = mouse_tp3,
                      .id = "time")

```


### Code chunk for your solutions

```{r}
#| label: challenge1solns
# create your function here

# implement your function 3 times

# bind the outputted data frames

```

:::


# Lists

## Lists, what are they?

-   Lists are a general purpose data structure in R. 
-   They are very flexible, in that they have *slots*. 
-   Unlike vectors, a `list` can be heterogeneous. 
    -   Each *slot* can contain elements of different types, such as vectors, matrices, data frames, or even other lists!
-   Lists allow you to organize and store diverse types of data in a single object.

## List basics

::: {.panel-tabset}
### Make a list

-   We'll start with a making a list and understand different properties about it.

```{r}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list
```

### List properties

-   Lists have a `length()`:

```{r}
length(my_list)

# note that there are no dimensions though:
dim(my_list)
```

-   `glimpse()` a list

```{r}
glimpse(my_list)
```

-   If the lists elements are named, you can ask for their names:

```{r}
names(my_list)
```


### Accessing list elements

::: {.panel-tabset}
#### Accessing with `$` 

-   We can access a single element in the list using `$` and its name, 
    -   similar to how we access data frame columns:

```{r}
names(my_list)

my_list$cat_names
my_list$hedgehog_names
my_list$dog_names
```

#### Accessing with `[[]]`

-   We can also access an element using the `[[]]` (double brackets) accessor 

```{r}
my_list[[1]]
my_list[[2]]
my_list[[3]]

# what happens when you run this?
# my_list[[4]]
```

-   If a list's slots are named, they can be accessed with their name:

```{r}
my_list[["hedgehog_names"]]

# what is the type/structure of the result?
str(my_list[["hedgehog_names"]])

str(my_list[["dog_names"]])

```
:::
:::

## Difference between `[]` and `[[]]`

-   Beware of the difference between `[]` and `[[]]`. 
-   The `[]` accessor returns a *list* of length 1, not the element in it:

```{r}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list["hedgehog_names"]
str(my_list["hedgehog_names"])

```

-   Compare to `[[]]`:

```{r}
my_list[["hedgehog_names"]]
```

In most cases, you want to use `[[]]`.


## Vectors vs. lists, and `[]` vs. `[[]]`

-   Vectors
    -   `vec[1:3]` returns a vector of length 3
    -   `vec[1]` returns a vector of length 1
    -   `[[]]` cannot be used with vectors

```{r}
my_vec <- c("Morris", "Julia", "Spiny", "Rover", "Spot")

my_vec[1:3]
my_vec[1]
```

-   Lists
    -   `list[1:3]` returns a list of length 3
    -   `list[1]` returns a list of length 1
    -   `list[[1]]` returns the first *element* of a list
        -   this could be a number (vector of length 1), a vector, another list, a data frame, a ggplot, whatever!

```{r}
my_list[1:3]
my_list[1]

my_list[[1]]

```

From the help manual of `[[`:

> "The most important distinction between `[`, `[[` and `$` is that the `[` can select more than one element whereas the other two select a single element."

## What does `my_list[[1]][1]` do?


```{r}
my_list[[1]][1]
```

-   Let's break it up step-by-step to better understand what it does:

```{r}
# save the first element of the list as tmpout
tmpout <-  my_list[[1]]
str(tmpout)

# now take the first element of tmpout
tmpout[1]
```

-   The first part, `my_list[[1]]`, returns the first element of `my_list` 
    -   which is a vector of length two (`c("Morris", "Julia")`) 
-   The second part, `[1]`, returns the first element of that vector "Morris":

```{r}
str(my_list[[1]][1])
```

-   If you try to pull out an element that doesn't exist, you get a strange variety of results, so be careful!!

```{r}
#| eval: false
# run in R on your computer to see the errors
my_list[[1]][3]

my_list[[6]]

my_list[["dogg_names"]] # check your spelling!!!
```

## Mini-challenge (your turn)

-   Find the length of the `cat_names` element in `my_list`. Then, find the length of `hedgehog_names`.

```{r}
#| label: list8
#| exercise: true
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))


```

## `purrr::pluck()`

-   The `purrr` package has an additional way to extract list elements: `purrr::pluck()`. 
-   This is equivalent to using `[[]]`, but always returns `NULL` when the element does not exist.
-   `pluck()` can be a handy way to extract something from a list without dealing with the `[[]]` and `[]` confusion.

```{r}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list %>%
  pluck("cat_names")

# go deeper in a list
my_list %>%
  pluck("cat_names", 2)
```

-   `pluck()` outputs a vector, just like `my_list[["cat_names"]]`:

```{r}
my_list %>%
  pluck("cat_names") %>%
  str()
```

-   `pluck()` outputs NULL instead of an error:

```{r}
my_list %>% 
  pluck("dogggg_names")

my_list %>% 
  pluck(10)
```



## Mini-challenge (your turn)

-   Run the code below. What does it output?

Did it output a `data.frame`, or a vector?

```{r}
penguins %>%
  pluck("species")
```

## Adding things to Lists

-   The *slots* in a `list` are not fixed. 
-   We can take an already made list and add a new slot to it using `<-`:

```{r}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list$raccoon_names <- c("Rocky")
my_list

my_list$cat_ages <- c(2, 5)
my_list
```

-   This is a little bit like adding a column to a data frame (with `mutate()` or `add_column()` or even `$`). 
  -   We can just keep adding!

-   This also works numerically:

```{r}
# another list, this time not named, but named 
my_list <- list(hello = c("hello", "world"),
                c("a","b","c"),
                stuff = "stuff")
# notice only two elements have names, this is allowed
my_list
names(my_list)

my_list[[4]] <- c("new","stuff")
my_list

my_list[[6]] <- "Does this work?"

# What happened to the fifth element?
my_list

names(my_list)
```

## *homogeneous* versus *heterogeneous* lists

::: {.panel-tabset}
### Difference

-   When we automate a repetitive process, we usually assume that the lists that we create are *homogeneous*. 
    -   That is, the data type of the list element is the same for each slot in the list. 
    -   In our case, usually the data.type will be a `data.frame`.

-   However, lists can also be *heterogenous* as we can see in the list above when we added `cat_ages` element.
-   A common pattern is to return a *heterogenous* list from a function. 
    -   For example, the `lm()` function actually returns a list.


### Example 1: `lm()` output

::: {.panel-tabset}
#### Run model

-   Run a linear regression model with lm()

```{r}
# outcome is on left side of ~
# predictor/independent variable/covariate  is on right side of ~
# specify data in data argument
output <- lm(body_mass_g ~ bill_length_mm, data = penguins)

# summary() produces summaries about the model results
sum_output <- summary(output)

sum_output
```

#### Output is a list

-    The summary regression output is a list:

```{r}
names(sum_output)
str(sum_output)
```

#### Access the coefficients

-   There are different ways to access the coefficients element of the list:

```{r}
sum_output$coefficients

sum_output[["coefficients"]]

sum_output %>% pluck(coefficients)
```

#### Object type of coefficients

-   Note this is a matrix, not a data frame!

```{r}
str(sum_output[["coefficients"]])
```


- What if we use `[]`?

```{r}
# [] returns a list:
sum_output["coefficients"]
str(sum_output["coefficients"])

```

-  Use `[]` to access multiple elements of the list:

```{r}
sum_output[c("call","coefficients")]

# a list of length 2
str(sum_output[c("call","coefficients")])

```

-   Remember you cannot use [[]] with more than one name:

```{r}
sum_output[[c("call","coefficients")]]
```

:::


### Example 3

-   This function runs a linear model using the penguins data and saves some of the regression output as a heterogeneous list

```{r}
run_penguin_model <- function(myisland = "Torgersen") {
  
  subset_penguins <- penguins %>%
    filter(island == myisland)
  
  output <- summary(lm(body_mass_g ~ bill_length_mm, data=subset_penguins))
  
  out_list <- list(
    coefficients = output$coefficients, 
    call = output$call, 
    island = myisland)
  
  return(out_list)
}

output_model <- run_penguin_model("Torgersen")

output_model
```
:::



## List of lists

-   Lists can be heterogeneous, and since they can contain anything, they can also contain other lists!

```{r}
my_info_list <- list(
  name = "Patty",
  pets = c("Cat", "Dog","Parrot","Kangaroo"),
  address = list(number = 273,
                 street = "N Main St",
                 city = "Portland",
                 state = "Oregon")
)

my_info_list

```

-   We can pull out elements from the list inside a list

```{r}
my_info_list[["address"]][["street"]]

my_info_list %>% pluck("address", "street")
```



## View a list

-   Try using `View()` with `my_list`, `my_info_list`, and `output` from the linear model above.
-   Rstudio tries to give you an interactive viewing experience for each element.
-   While in the View screen, click on the icon at the very right of the screen, what does that do?


```{r}
#| eval: false
# run this in R

View(my_list)

View(my_info_list)

View(output)
```



## `data.frame`s are list-like

-   You can access columns in a `data.frame` using `$`. 
-   This is because a `data.frame` is a special *instance* of a list.
-   Hence, you can use all of the above accessors to manipulate variables in a `data.frame`.

-   Load the `mtcars` data frame (it's nice and small)

```{r}
data(mtcars)
glimpse(mtcars)

```

-   What is `length` doing here? how does it compare to `dim`?

```{r}
length(mtcars)
```

-   `mtcars` is also a kind of list - we can use `$, [[]],` and `pluck()`:

```{r}
mtcars$mpg
mtcars[[1]]
str(mtcars[[1]])

mtcars[["mpg"]]

mtcars %>% pluck("mpg")

mtcars %>% pull("mpg")
```

-   Compare the output above to that below when using a single `[]`:

```{r}
mtcars[1]
str(mtcars[1])
```


-   A common pattern in Base-R for filtering that you should be aware of is using the `$` operator for filtering on rows. -   The `tidyverse` simplifies this with `filter()` and that's what I encourage you to use, 
    -   but this pattern is often used with objects in the Bioconductor package (or in other people's code ;))

```{r}
mtcars[mtcars$cyl == 8, ]
```

-   This is the tidyverse version of the above "base-R" code

```{r}
mtcars %>% 
  filter(cyl == 8)
```


## Recap of lists

-   These are some basic operations with lists in R. 
-   Lists are powerful for organizing and manipulating complex data structures in a flexible way.


::: {.panel-tabset}
### Creating a List:

-   You can create a list using the `list()` function. 
-   Elements within the list can be of different types:

```{r}
# Example list
my_list <- list(
  numeric_vector = c(1, 2, 3),
  character_vector = c("a", "b", "c"),
  matrix = matrix(1:4, nrow = 2),
  data_frame = data.frame(name = c("John", "Jane"), age = c(25, 30))
)

my_list
```

### Accessing Elements:

-   You can access elements in a list using 
    -   double square brackets `[[]]` 
    -   or the dollar sign `$` for named elements:

```{r}
# Accessing elements
(numeric_vector <- my_list[[1]])
(character_vector <- my_list$character_vector)
```

### Adding Elements:

-   You can add elements to a list 
    -   using the `c()` function or 
    -   by assigning a new element to a specific index or 
    -   using the familiar `$`:

```{r}
# Adding elements
my_list <- c(my_list, new_element = "This is a new element")

my_list[[6]] <- "Another new element"
my_list$new3 <- "Yet another new element"

my_list
```

### List Attributes:

-   Lists can have attributes, such as names and dimensions. 
-   You can assign names to list elements:

```{r}
names(my_list)

# Assigning names to list elements
names(my_list) <- c("numeric", "character", "matrix", "dataframe", "new1", "new2", "new3")

names(my_list)
```

### Nested Lists:

-   Lists can be nested, meaning you can have lists within lists:

```{r}
# Nested list
nested_list <- list(
  sublist1 = list(a = 1, b = 2),
  sublist2 = list(c = "apple", d = "banana")
)

nested_list
```



:::



# Next time

Next class we will focus on iteration using functions, lists, and `purrr::map()`.



# Post Class Survey

Please fill out the [post-class survey](https://ohsu.ca1.qualtrics.com/jfe/form/SV_2f0SKdTYUo80aRo){target="_blank"}. 

Your responses are anonymous in that I separate your names from the survey answers before compiling/reading. 

You may want to review previous years' feedback [here](https://niederhausen.github.io/BSTA_526_W26/survey_feedback_previous_years){target="_blank"}.



# Acknowledgements

-   Part 8 is based on the BSTA 505 Winter 2023 course, taught by Jessica Minnier.
    -   I made modifications to update the material from RMarkdown to Quarto, and streamlined/edited content for slides.

-   Minnier's Acknowledgements:
    -   Written by Jessica Minnier and Ted Laderas. 
    -   Some material inspired by [Kelly Bodwin's Adventures in R](https://www.adventures-in-r.com/post/05-loops-packages/).


