---
title: "`details::details`"
pagetitle: "Functions of the Week"
subtitle: "BSTA 526 Functions of the Week"
description: "Collapsible sections in quarto"
author: "Walter Ell"
date: "3/5/2026"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

#| label: setup
#| include: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

library(palmerpenguins)

```

See the [syllabus](https://niederhausen.github.io/BSTA_526_W26/syllabus.html#function-of-the-week) for links to presentations from previous years.



# `Function(s) Name(s)`

> What function(s) is being presented and what package is it from?

```{r}
#::: {.callout-note collapse="true"}
# ## Two # is one way to make the title
# Content here
#:::

#::: {.callout-note collapse="true" title=" Different title"}
# Content here
#:::


```



# What is it for?

::: {.callout-note collapse="true"}
## For content labeled for notes

::: {.callout-tip collapse="true"}
## For content used for tips

Placing text within the nested layer above can cause problems
:::
:::


This formatting allows for quarto to format with collapsible sections to reduce visual clutter.

# Examples

::: {.callout-caution collapse="true"}
## For content used for cautionary warnings

The three : that are used for the wrapper can cause issues with some packages
when using Python
:::

::: {.callout-important collapse="true"}
## For content used for important info

:::

::: {.callout-tip collapse="true" title="They can have different titles"}
## Be care ful about changing the title

It often overrides other formats within
:::


::: {.callout-caution collapse="true" title="R code can be placed within"}
```{r}
head(penguins)
```
:::

```{r}
#::: {.callout-important collapse="true" icon=false appearance=minimal}
# Content here
#:::

```

::: {.callout-important collapse="true" icon=false appearance=minimal}
## Customization
The icon can be turned off and the appearance changed
:::



# Is it helpful?

Useful for making the vertical space of a quarto document less cluttered and more selective.
However the formating can be disrupted by other quarto formating. Some extentions
to the customization can be done via terminal commands.
