---
title: "HW 1"
subtitle: "BSTA 526"
author: "Your name here - update this!!!!"
date: "Due 1/16/25"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    theme: cosmo
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions for Submitting

Submit your knitted html file - if you are having issues, you can submit the `.qmd` file, put please do your best to render it. 

Make sure to name your assignment "hw_01_b526_LASTNAME.html" and to fill out the author field with your name. 

**Due Date: January 16, 2025; 11:55 pm via Sakai**



# Assignment

## 1. Object manipulation (30%)

Create an object called `agge` that contains your age in years (or perhaps the age you wish you were). Then reassign the object to a new object called `age` (e.g., correct the typo). Then remove the previous object from your environment and then calculate your age in days

```{r}
#create agge object

#reassign agge value to age

#remove agge object

#Use math to calculate your age in days

```

## 2. Vector manipulation (character data) (30%):

Create a object called `buildings` representing a vector that contains four names of buildings on OHSU's campus, including the building where you work/study (here's a reference: <https://www.ohsu.edu/visit/maps>). If you work from home, add element `home` to the vector.

Add `Portland, Oregon` to the beginning of the vector, and `Phys Plant` to the end of the vector

subset the vector to show only the building in which you work/study.

```{r}
#create buildings object
buildings <- c()

#Add "Portland, Oregon" to the beginning of buildings

#Add "Phys Plant" to the end of buildings

#subset buildings to show only your building
buildings[]

```

## 3. Vector manipulation (numerical data) (20%):

The following vector represents the number of vacation days used by various employees.

How many employees are represented in the vector?

How many vacation days have been used in total?

```{r}
vacation_days <- c(5, 7, 20, 1, 0, 0, 12, 4, 2, 2, 2, 4, 5, 6, 7, 10, 4)

#how many employees are represented in the vector?

#how many vacation days total?

```


## 4. Vectorization (20%)


Create a logical vector that represents whether the number of vacation days taken by each employee is less than 3. Then, add together the entire vector to determine how many employees have taken less than 3 vacation days.

```{r}
vacation_logical <-  c() # replace c() with your code
  
# use class() to check this is a logical vector
class(vacation_logical)

# how many employees have taken less than 3 vacation days
```

